---
title: "World Cup 2022"
output:
  html_notebook: default
  pdf_document: default
---

```{r, echo=FALSE}
library(rlist)
library(dplyr)
library(skimr)
library(plotly)
library(scales)
library(corrplot)
library(tidyverse)
library(InformationValue)
```

The world cup 2022 will be hosted by Qatar in 2022. We wanted to try and predict
which country has the biggest probability of winning the tournament by using supervised learning.

# 1. Data

The data that we will be using is already cleaned and does not require pre-processing.

The data set provides a complete overview of all international football matches played since the 90s. On top of that, the strength of each team is provided by incorporating the FIFA rankings as well as player strengths based on the EA Sport FIFA video game. It is available on [kaggle](https://www.kaggle.com/datasets/brenda89/fifa-world-cup-2022)

## Variables

- `date` : date of the match
- `home_team` : name of the home team
- `away_team` : name of the away team
- `home_team_continent` : continent of the home team
- `away_team_continent` : continent of the away team
- `home_team_fifa_rank` : FIFA rank of the home team when the match took place
- `away_team_fifa_rank` : FIFA rank of the away team when the match took place
- `home_team_total_fifa_points` : total number of FIFA points of the home team at the time of the match
- `away_team_total_fifa_points` : total number of FIFA points of the away team at the time of the match
- `home_team_score` : full-time home score (excluding penalty shootout)
- `away_team_score` : full-time away score (excluding penalty shootout)
- `tournament` : name of tournament
- `city` : name of the city where the match was played
- `country` : name of the country where the match was played
- `neutral_location` :
  - `TRUE` : the match was played at a neutral venue
- `shoot_out`:
  - `TRUE` : the match included a penalty shootout
- `home_team_result` : result of the home team (including penalty shootout)
- `home_team_goalkeeper_score` : FIFA game score of the highest ranked GK of the home team
- `away_team_goalkeeper_score` : FIFA game score of the highest ranked GK of the away team
- `home_team_mean_defense_score` : Average FIFA game score of the 4 highest ranked defensive players of the home team
- `away_team_mean_defense_score` : Average FIFA game score of the 4 highest ranked 
defensive players of the away team
- `home_team_mean_midfield_score` : Average FIFA game score of the 4 highest ranked midfield players of the home team
- `away_team_mean_midfield_score` : Average FIFA game score of the 4 highest ranked midfield players of the away team
- `home_team_mean_offense_score` : Average FIFA game score of the 3 highest ranked attacking players of the home team, including wing players
- `away_team_mean_offense_score` : Average FIFA game score of the 3 highest ranked attacking players of the away team, including wing players

# 2. EDA

```{r}
# class of each variable
spec(input_data)
```
```{r}
# summary
skim_without_charts(input_data)
```

## Missing data

```{r}
input_data %>%
  summarise_all(list(~is.na(.)))%>%
  pivot_longer(everything(),
               names_to = "variables", values_to="missing") %>%
  count(variables, missing) %>%
  ggplot(aes(y=variables,x=n,fill=missing))+
  geom_col()+
  scale_fill_manual(values=c("#A3BE8C","#EBCB8B"))+
  theme(axis.title.y=element_blank())
```


## Top 10 teams in 2022

```{r}
# Get the ranking of all home teams
home <-
  input_data %>% 
  select(date, home_team, home_team_fifa_rank) %>% 
  rename(team = home_team, ranking = home_team_fifa_rank)

# Get the ranking of all away teams
away <-
  input_data %>% 
  select(date, away_team, away_team_fifa_rank) %>% 
  rename(team = away_team, ranking = away_team_fifa_rank)

# Combine both data frames into one
fifa_ranking <- rbind(home, away)

# Get the latest ranking of each country based on their most recent match
latest_fifa_ranking <-
  fifa_ranking %>% 
  arrange(team, desc(date)) %>% 
  group_by(team) %>% 
  mutate(row_number = row_number(team)) %>% 
  filter(row_number == 1) %>% 
  select(-row_number, -date) %>% 
  arrange(ranking)
  
head(latest_fifa_ranking, 10)
```

## FIFA rankings over time

```{r}
top5_list <- head(latest_fifa_ranking, 5)$team

top5_ranking <-
  fifa_ranking  %>% 
  filter(team %in% top5_list)

p <-
  ggplot(data = top5_ranking,
         mapping = aes(
           x = date,
           y = ranking,
           group = team,
           color = team
         )) +
  geom_line() +
  scale_y_reverse() +
  labs(
    x = "Date",
    y = "FIFA Ranking",
    color = "Team",
    title = "FIFA Rankings of the 2022 Top 5 teams"
  )

ggplotly(p)

```

## Teams with strongest GK

```{r}
# Gather goalkeeper data from matches
gk_home <-
  input_data %>% 
  select(date, home_team, home_team_goalkeeper_score) %>% 
  rename(team = home_team, goalkeeper_rating = home_team_goalkeeper_score)

gk_away <-
  input_data %>% 
  select(date, away_team, away_team_goalkeeper_score) %>% 
  rename(team = away_team, goalkeeper_rating = away_team_goalkeeper_score)

gk_rating <- drop_na(rbind(gk_home, gk_away))

# Get latest rating of each team's goalkeeper and show top 10
latest_gk_rating <-
  gk_rating %>% 
  arrange(team, desc(date)) %>% 
  group_by(team) %>% 
  mutate(row_number = row_number(team)) %>% 
  filter(row_number == 1) %>% 
  select(-row_number, -date) %>% 
  arrange(-goalkeeper_rating)

ggplot(data = head(latest_gk_rating, 10), mapping = aes(x=goalkeeper_rating, y=reorder(team, goalkeeper_rating), label=goalkeeper_rating)) +
  geom_col(fill="#88C0D0") +
  geom_text(position = position_stack(vjust = 0.5)) +
  labs(title = "Top 10 teams with the strongest goalkeeper",
       subtitle = "Based on the highest rated goalkeeper of each team",
       x="Goalkeeper Rating",
       y="Country")
```
### Teams with strongest defense

```{r}
# Gather goalkeeper and defense data from matches
def_home <-
  input_data %>% 
  select(date, home_team, home_team_goalkeeper_score, home_team_mean_defense_score) %>% 
  rename(team = home_team, goalkeeper_rating = home_team_goalkeeper_score, mean_defense_rating = home_team_mean_defense_score)

def_away <-
  input_data %>% 
  select(date, away_team, away_team_goalkeeper_score, away_team_mean_defense_score) %>% 
  rename(team = away_team, goalkeeper_rating = away_team_goalkeeper_score, mean_defense_rating = away_team_mean_defense_score)

def_rating <- drop_na(rbind(def_home, def_away))

# Get latest combined rating of each team and show top 10
latest_def_rating <-
  def_rating %>% 
  arrange(team, desc(date)) %>% 
  mutate(total_def = goalkeeper_rating + mean_defense_rating) %>% 
  group_by(team) %>% 
  mutate(row_number = row_number(team)) %>% 
  filter(row_number==1) %>% 
  arrange(-total_def) %>% 
  select(-row_number, -date)

ggplot(data = head(latest_def_rating, 10), mapping=aes(x=total_def, y=reorder(team, total_def), label=total_def)) + 
  geom_col(fill="#88C0D0") +
  geom_text(position = position_stack(vjust = 0.5)) +
  labs(title = "Top 10 teams with the strongest defense",
       subtitle = "Based on goalkeeper and mean defense ratings",
       x = "Total Defense Rating",
       y = "Teams") 
```

### Teams with strongest midfield

```{r}
mid_home <-
  input_data %>% 
  select(date, home_team, home_team_mean_midfield_score) %>% 
  rename(team = home_team, midfield_rating = home_team_mean_midfield_score)

mid_away <-
  input_data %>% 
  select(date, away_team, away_team_mean_midfield_score) %>% 
  rename(team = away_team, midfield_rating = away_team_mean_midfield_score)

mid_rating <- drop_na(rbind(mid_home, mid_away))

# Get latest midfield rating of each team and show top 10
latest_mid_rating <-
  mid_rating %>% 
  arrange(team, desc(date)) %>% 
  group_by(team) %>% 
  mutate(row_number = row_number(team)) %>% 
  filter(row_number == 1) %>% 
  arrange(-midfield_rating) %>% 
  select(-date, -row_number)

ggplot(data = head(latest_mid_rating, 10), mapping=aes(x=midfield_rating, y=reorder(team, midfield_rating), label=midfield_rating)) + 
  geom_col(fill= "#88C0D0") +
  geom_text(position = position_stack(vjust = 0.5)) +
  labs(title = "Top 10 teams with the strongest midfield",
       subtitle = "Based on the average rating of the 4 highest rated midfield players of each team",
       x = "Midfield Rating",
       y = "Teams")
```

### Teams with strongest offense

```{r}
off_home <-
  input_data %>% 
  select(date, home_team, home_team_mean_offense_score) %>% 
  rename(team = home_team, offense_rating = home_team_mean_offense_score)

off_away <-
  input_data %>% 
  select(date, away_team, away_team_mean_offense_score) %>% 
  rename(team = away_team, offense_rating = away_team_mean_offense_score)

off_rating <- drop_na(rbind(off_home, off_away))

# Get latest offense rating of each team and show top 10
latest_off_rating <-
  off_rating %>% 
  arrange(team, desc(date)) %>% 
  group_by(team) %>% 
  mutate(row_number = row_number(team)) %>% 
  filter(row_number == 1) %>% 
  arrange(-offense_rating) %>% 
  select(-date, -row_number)

ggplot(data = head(latest_off_rating, 10), mapping=aes(x=offense_rating, y=reorder(team, offense_rating), label=offense_rating)) +
  geom_col(fill="#88C0D0") +
  geom_text(position = position_stack(vjust = 0.5)) +
  labs(title="Top 10 teams with the strongest offense",
      subtitle="Based on the average rating of the 3 highest rated offensive players of each team",
      x="Offense Rating",
      y="Teams")
```

## Is it better to play at home ?

```{r}
home_team_advantage <-
  input_data %>% 
  filter(neutral_location == FALSE) %>% 
  count(home_team_result) %>% 
  mutate(percentage = label_percent()(n/sum(n)))

ggplot(data = home_team_advantage, mapping=aes(x="", y=n, fill=home_team_result)) +
  geom_bar(width = 1, stat = "identity", color="white") +
  geom_text(aes(label=paste0(percentage)), position= position_stack(vjust=0.5)) +
  coord_polar("y") +
  scale_fill_manual(values = c("#EBCB8B", "#BF616A",
                               "#A3BE8C")) +
  theme_void() +
  labs(title = "Distribution of match results of home teams",
       subtitle = "Excluding matches played at neutral locations",
       fill="Result")
```

## Correlation Matrix

```{r}
# select numeric columns only
input_numeric_data <- input_data %>%
  select_if(is.numeric) %>%
  drop_na()

# rename variables for easier correlation plot visualization
input_numeric_data <- input_numeric_data %>% rename(
  rank1 = home_team_fifa_rank,
  rank2 = away_team_fifa_rank,
  total_fifa_points1 = home_team_total_fifa_points,
  total_fifa_points2 = away_team_total_fifa_points,
  score1 = home_team_score,
  score2 = away_team_score,
  gk_score1 = home_team_goalkeeper_score,
  gk_score2 = away_team_goalkeeper_score,
  df_score1 = home_team_mean_defense_score,
  df_score2 = away_team_mean_defense_score,
  att_score1 = home_team_mean_offense_score,
  att_score2 = away_team_mean_offense_score,
  mf_score1 = home_team_mean_midfield_score,
  mf_score2 = away_team_mean_midfield_score
)

# create correlation plot
input_numeric_data %>%
  cor() %>%
  corrplot(
    type = "upper",
    diag = FALSE,
    col=colorRampPalette(c("firebrick","lightyellow","green4"))(100),
    method = "shade",
    shade.col = NA,
    tl.col = "black",
    tl.srt = 45
  )
```

# 2. Data Processing / Feature Engineering

## Create new features

```{python}
import numpy as np
import pandas as pd


df = r.input_data
df['date'] = pd.to_datetime(df['date'])

df['rank_difference'] = df['home_team_fifa_rank'] - df['away_team_fifa_rank']
df['average_rank'] = (df['home_team_fifa_rank'] + df['away_team_fifa_rank'])/2
df['point_difference'] = df['home_team_total_fifa_points'] - df['away_team_total_fifa_points']
df['score_difference'] = df['home_team_score'] - df['away_team_score']
df['is_won'] = df['score_difference'] > 0 # take draw as lost
df['is_stake'] = df['tournament'] != 'Friendly'
```

## Model

```{python, echo = FALSE, message = FALSE, results='hide'}
#!pip install matplotlib
#!pip install seaborn
#!pip install datetime

import seaborn as sns
import matplotlib.pyplot as plt

from sklearn import linear_model
from sklearn import ensemble
from sklearn.model_selection import train_test_split
from sklearn.metrics import confusion_matrix, roc_curve, roc_auc_score
from sklearn.pipeline import Pipeline
from sklearn.preprocessing import PolynomialFeatures



X, y = df.loc[:,['average_rank', 'rank_difference', 'point_difference']], df['is_won']
X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=0.2, random_state=42)

logreg = linear_model.LogisticRegression(C=1e-5)
features = PolynomialFeatures(degree=2)
model = Pipeline([
    ('polynomial_features', features),
    ('logistic_regression', logreg)
])
model = model.fit(X_train, y_train)

# figures 
fpr, tpr, _ = roc_curve(y_test, model.predict_proba(X_test)[:,1])
plt.figure(figsize=(15,5))
ax = plt.subplot(1,3,1)
ax.plot([0, 1], [0, 1], 'k--')
ax.plot(fpr, tpr)
ax.set_title('AUC score is {0:0.2}'.format(roc_auc_score(y_test, model.predict_proba(X_test)[:,1])))
ax.set_aspect(1)

ax = plt.subplot(1,3,2)
cm = confusion_matrix(y_test, model.predict(X_test))
ax.imshow(cm, cmap='Blues', clim = (0, cm.max())) 

ax.set_xlabel('Predicted label')
ax.set_title('Test set')

ax = plt.subplot(1,3,3)
cm = confusion_matrix(y_train, model.predict(X_train))
ax.imshow(cm, cmap='Blues', clim = (0, cm.max())) 
ax.set_xlabel('Predicted label')
ax.set_title('Training set')


plt.show()

```

## Simulation

```{python}

#List of all Teams in 2022 World Cup
teams_worldcup = ['Qatar', 'Ecuador', 'Senegal', 'Netherlands', 'England', 'Iran', 'USA',
                  'Wales', 'Argentina', 'Saudi Arabia', 'Mexico', 'Poland', 'France', 
                  'Australia', 'Denmark', 'Tunisia', 'Spain', 'Costa Rica', 'Germany', 
                  'Japan', 'Belgium', 'Canada', 'Morocco', 'Croatia', 'Brazil', 'Serbia', 
                  'Switzerland', 'Cameroon', 'Portugal', 'Ghana', 'Uruguay', 'South Korea']

```

```{python}

world_cup_rankings_home = df[['home_team','home_team_fifa_rank','home_team_total_fifa_points']].loc[df['home_team'].isin(teams_worldcup) & (df['date']>'2021-01-01')]
world_cup_rankings_away = df[['away_team','away_team_fifa_rank','away_team_total_fifa_points']].loc[df['away_team'].isin(teams_worldcup)& (df['date']>'2021-01-01')]
world_cup_rankings_home = world_cup_rankings_home.set_index(['home_team'])


```

```{python}
world_cup_rankings_home = world_cup_rankings_home.groupby('home_team').mean()
world_cup_rankings_away = world_cup_rankings_away.groupby('away_team').mean()
```
```{python}
    candidates = [ 'Senegal', 'Netherlands', 'England','USA', 'Argentina', 'Poland', 'France', 'Denmark','Spain', 'Germany', 'Belgium', 'Croatia', 'Brazil', 'Serbia', 'Portugal','Uruguay']
    finals = ['round_of_16', 'quarterfinal', 'semifinal', 'final']
```



```{python}
from tqdm import tqdm

simulation_winners = list()
simulation_results_winners = list()
simulation_results_round16 = list()
simulation_df_round16 = list()
simulation_results_quarterfinal = list()
simulation_df_quarterfinal = list()
simulation_results_semifinal = list()
simulation_df_semifinal = list()

n_simulations = 1000

for j in tqdm(range(n_simulations)):
  
    for f in finals:
        iterations = int(len(candidates) / 2)
        winners = []
        prob = []
        
        for i in range(iterations):
            home = candidates[i*2]
            away = candidates[i*2+1]
            row = pd.DataFrame(np.array([[np.nan, np.nan, True]]), columns=X_test.columns)
            home_rank = world_cup_rankings_home.loc[home, 'home_team_fifa_rank']
            home_points = world_cup_rankings_home.loc[home, 'home_team_total_fifa_points']
            opp_rank = world_cup_rankings_away.loc[away, 'away_team_fifa_rank']
            opp_points = world_cup_rankings_away.loc[away, 'away_team_total_fifa_points']
            row['average_rank'] = (home_rank + opp_rank) / 2
            row['rank_difference'] = home_rank - opp_rank
            row['point_difference'] = home_points - opp_points
            home_win_prob = model.predict_proba(row)[:,1][0]
            simulated_outcome = np.random.binomial(1, home_win_prob)
            winners.append(away) if simulated_outcome <= 0.5 else winners.append(home)
            prob.append(1 - simulated_outcome) if simulated_outcome <= 0.5 else prob.append(simulated_outcome)
            
        if f == 'round_of_16':
            step_16 = ['round_16'] * 8
            candidates_round_16 = zip(step_16, winners, prob)
            df_candidates_round_16 = pd.DataFrame(candidates_round_16, columns = ['Step','Team','Prob'])
            simulation_df_round16.append(df_candidates_round_16)
            simulation_results_round16.append(winners)
            
        if f == 'quarterfinal':
            step_quarterfinal = ['quarterfinal'] * 4
            candidates_quarterfinal = zip(step_quarterfinal,winners, prob)
            df_candidates_quarterfinal = pd.DataFrame(candidates_quarterfinal, columns = ['Step','Team','Prob'])
            simulation_df_quarterfinal.append(df_candidates_quarterfinal)
            simulation_results_quarterfinal.append(winners)
            
        if f == 'semifinal':    
            step_semifinal = ['semifinal'] * 2
            candidates_semifinal = zip(step_semifinal,winners, prob)
            df_candidates_semifinal = pd.DataFrame(candidates_semifinal, columns = ['Step','Team','Prob'])
            simulation_df_semifinal.append(df_candidates_semifinal)
            simulation_results_semifinal.append(winners)
            
        if f == 'final':    
            step_final = ['final'] * 1
            candidates_final = zip(step_final,winners, prob)
            df_candidates_final = pd.DataFrame(candidates_final, columns = ['Step','Team','Prob'])
            simulation_winners.append(df_candidates_final)
            simulation_results_winners.append(winners)
            
        list_win = list(zip(winners,prob))
        candidates = winners 
df_candidates_round_16 = pd.concat(simulation_df_round16)
df_candidates_quarterfinal = pd.concat(simulation_df_quarterfinal)
df_candidates_semifinal = pd.concat(simulation_df_semifinal)
df_candidates_final = pd.concat(simulation_winners)
df_results = pd.concat([df_candidates_round_16,df_candidates_quarterfinal,df_candidates_semifinal,df_candidates_final]) #final DataFrame
simulation_results_round16 = sum(simulation_results_round16, [])
simulation_results_quarterfinal = sum(simulation_results_quarterfinal, [])
simulation_results_semifinal = sum(simulation_results_semifinal, [])
simulations_winners = sum(simulation_results_winners, [])
```



```{python}
for j in tqdm(range(n_simulations)):
  print(j+1)
```












